Fail Faster

### 2025-03-20 paper/paper.tex
non-sequitur removal

Of course, simply arguing that the randomness library is on the hot
path does not guarantee that a faster sampling leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the randomness library for \bq{} with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment:
OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much faster bugs can be found just by using a more performant randomness library.

\section{Speeding up Generator DSLs by Staging and Faster Randomness Libraries}
\label{section:faster-generators}
\jwc{This section needs a much better title...}
% \jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}

@@

In Muli-stage programming (also known more simply as staging), programs execute
over the course of multiple stages, with each stage producing the code to be run
in the next.  This is an old idea in programming languages, with roots going
back to quasiquotation in LISP \cn{}.  For the purposes of this paper, we will
primarily consider two stages, namely compile time and run time. Staged programs thus
execute twice: once at compile time, which produces more code, which is then
compiled and run at run time.

@@

\jwc{Unclear what we should call this section}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another bottleneck for finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \jwc{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the natural experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow randomness library with a faster but extensionally equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.

>>>>

Of course, simply arguing that the randomness library is on the hot
path does not guarantee that a faster sampling leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment:
the \rand{} that \bq{} uses --- OCaml's \texttt{Splittable\_random} library --- is slow in a way that
can be improved \emph{without} changing its behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much faster bugs can be found just by using a more performant randomness library.

\section{\name{}}
\label{section:faster-generators}
\jwc{This section needs a much better title...}
% \jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}

@@

In Muli-stage programming (also known more simply as staging), programs execute
over the course of multiple stages, with each stage producing the code to be run
in the next. For the purposes of this paper, we will
primarily consider two stages, namely compile time and run time. Staged programs thus
execute twice: once at compile time, which produces more code, which is then
compiled and run at run time.

@@

\jwc{Unclear what we should call this section}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another bottleneck for finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \jwc{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the natural experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow randomness library with a faster but semantically equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.
